Boosting RAG Efficiency with dyRAG: Dynamic Candidate Selection for Optimal Retrieval

### 2025-07-18 Paper/Submission3/RAG3.tex
All changes
Last submitted paper is in folder submission3

%% This is a JAIR Example File Compiled by Nicholas Mattei (nsmattei@tulane.edu) 
%% and Odd Erik Gundersen (odderik@ntnu.no)
%% and Mykel Kochenderfer (mykel@stanford.edu)
%% March 29, 2025
%%
%% This file is based off the ACM Latex Template https://www.acm.org/publications/proceedings-template
%% Revision 2.12 (12/28/2024)
%% 
%% Please see https://www.jair.org/index.php/jair for more information and submission instructions.
%%

%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{jair}.
%%
\documentclass[review]{jair}

\setcopyright{cc}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{10.1613/jair.1.xxxxx}

%%
\JAIRAE{Insert JAIR AE Name}
\JAIRTrack{Insert JAIR Track Name Here}
\acmVolume{4}
\acmArticle{111}
\acmMonth{8}
\acmYear{2025}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\usepackage{graphicx}%
\usepackage{multirow}%
%\usepackage{amsfonts, amsmath,amssymb}%
%%\usepackage{amsthm}%
%%\usepackage{mathrsfs}%
%\usepackage[title]{appendix}%
%\usepackage{xcolor}%
%\usepackage{textcomp}%
%\usepackage{manyfoot}%
%\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{subcaption}
\usepackage{booktabs} % For professional-quality rules
\usepackage{makecell} % For multiline headers
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
%\title{JAIR Example Template}
\title[dyRAG.Dynamic K-Selection]{Boosting RAG Efficiency with dyRAG: Dynamic K-Selection for Optimal Retrieval}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research and/or corresponding author.
\author{Zoubida Asmaa Boudjenane}

\email{asmaa.boudj04@gmail.com}
\orcid{0000-0002-2037-3694}
% \orcid{https://orcid.org/0000-0002-2037-3694} % This works too if you want to see the full URL
\affiliation{%
  \institution{Computer Science Department, University of Mascara}
  \city{Mascara}
  \state{Mascara}
  \country{Algeria}
}

\author{Mohammed Salem}
\orcid{0000-0002-2037-3694}
\authornote{Corresponding Author.}
\email{salem@univ-mascara.dz}
\affiliation{%
  \institution{LISYS Laboratory, University of Mascara}
  \city{Mascara}
  \state{Mascara}
  \country{Algeria}}

%\author{Holger H. Hoos}
%\orcid{0000-0003-0629-0099}
%\email{hoos@cs.ubc.ca}
%\affiliation{%
%  \institution{University of British Columbia}
%  \city{Vancouver}
%  \state{British Columbia}
%  \country{Canada}
%}

%\author{Kevin Leyton-Brown}
%\orcid{0000-0002-7644-5327}
%\email{kevinlb@cs.ubc.ca}
%\affiliation{%
%  \institution{University of British Columbia}
%  \city{Vancouver}
%  \state{British Columbia}
%  \country{Canada}
%}

%\author{Ben Trovato}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Boudjenane and Salem.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
{\bf Background:} 
    Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge. However, its effectiveness heavily depends on the quality of the retrieved documents. Using a fixed number of retrieved documents (K) may fail to adapt to varying query complexity, often leading to suboptimal retrieval—either including irrelevant documents or missing crucial ones.

    {\bf Objectives:}
   To address this issue, we propose dyRAG, a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency. Our goal is to improve retrieval performance by maximizing relevant information and minimizing noise.

    {\bf Methods:}
   We introduce dyRAG, which dynamically adjusts the number of retrieved documents (K) in response to query complexity and dataset characteristics. The method balances precision and computational efficiency to optimize RAG systems.

    {\bf Results:}
    Experimental results show that dyRAG outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems. It demonstrates greater adaptability across various contexts compared to traditional methods.

    {\bf Conclusions:}
    dyRAG provides a dynamic and efficient approach to document retrieval in RAG systems, improving performance by tailoring retrieval to query-specific needs. This advancement enhances the flexibility and effectiveness of RAG in diverse applications.
\end{abstract}

%\begin{abstract}
%      A clear and well-documented \LaTeX\ document is presented as an
%  article formatted for publication by ACM in a conference proceedings
%  or journal publication. Based on the ``acmart'' document class, this
%  article presents and explains many of the common variations, as well
%  as many of the formatting elements an author may use in the
%  preparation of the documentation of their work.
%\end{abstract}

%% JAIR Note: 
%% Do not include ACM CCS Concepts or Keywords

%% To be updated by authors.
\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text\cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}. Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}. Despite their impressive capabilities, LLMs suffer from hallucination\cite{Radford2019}, often generating confident but incorrect information. This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{Radford2019}. Additionally, their static nature being trained on fixed datasets\cite{Radford2019} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.

Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}, fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks. Despite their strengths, they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{Chuyuan2025}, which integrates real-time retrieval from external knowledge sources with generative models, enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations. For instance, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally, a critical challenge is determining the optimal number of retrieved passages K to feed into the generator\cite{lewis2020retrieval}. Often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes K a key factor in balancing performance and efficiency.

@@

In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs. The use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches. Unlike traditional approaches, dyRAG adapts to the query and dataset characteristics, making it more flexible for diverse contexts. Furthermore, its design for handling large datasets ensures scalable and precise retrieval, which is crucial for RAG systems working with extensive corpora.

@@

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method, the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001}, TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents K is fixed, it fails to adapt to the varying complexity of queries. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.

Dynamic top-K Selection in Retrieval :addresses the limitations of static K by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal K for retrieval, a reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust K by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of K Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations. They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability. These systems might use a static K as a baseline but dynamically adjust it such as, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} First, it retrieves an initial set of $k_1$ documents. Then, a classifier dynamically evaluates their relevance and retrieves additional relevant $K_2$ documents to enhance recall. However hybrid methods may face challenges in determining when to apply static or dynamic K-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt K to the unique characteristics of each query and dataset.

@@

\section{Overview of Retrieval-Augmented Generation }\label{sec3}
Retrieval-Augmented Generation (RAG) is a method designed to enhance the performance of large language models (LLMs) by integrating external, reliable knowledge sources into their response generation process\cite{awsRAG}. RAG combines search capabilities with LLM prompting. The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt\cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.JPG}
	\caption{RAG system component\cite{gupta20241}. }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components: the retriever and the generator. The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows :
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}which relies on exact term matching techniques like BM25 and TF-IDF, dense retrieval\cite{karpukhin2020dpr}which Uses deep learning models to encode queries and documents into dense vector embeddings, retrieving relevant documents based on semantic similarity rather than exact word matching, and others. For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
1. Encoding each object into a specific representation (e.g., term-based for sparse retrieval, vector embeddings for dense retrieval).\\
2. Building an index to organize and efficiently retrieve relevant data during search .

This structured approach ensures that the most relevant information is retrieved quickly and accurately\cite{zhao2024retrievalaugmentedgenerationaigeneratedcontent}.

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response.

@@

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation, Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images, for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

@@

\begin{table}[h]
	\centering
	\large  % Increase overall font size
	\begin{tabular}{|c|p{10cm}|}  % Use p{width} for text wrapping
		\hline
		\textbf{Notation} & \textbf{Description} \\ \hline
		\( q \) (\( Q \), \( |Q| \)) & A single query (set of all queries, total number of queries). \\ \hline
		\( x \) (\( X \), \( |X| \)) & A single item (set of all items, total number of items). \\ \hline
		\( \phi(q, x) \) & The learned similarity function, Mixture-of-Logits (MoL), which computes the similarity between \( q \) and \( x \). \\ \hline
		\( P \) (\( P_q \), \( P_x \)) & Total number of low-rank embeddings (\( P = P_q \times P_x \)), where \( P_q \) and \( P_x \) are the number of embeddings for \( q \) and \( x \), respectively. \\ \hline
		\( \pi_p(q, x) \)  & Weight for the \( p \)-th (or \( p_q \)-th and \( p_x \)-th) embedding set for \( (q, x) \). \\ \hline
		\( f(q) \) (\( f_p(q) \)) & Learned embedding for the query (\( p \)-th component-level embedding for \( q \)). \\ \hline
		\( g(x) \) (\( g_p(x) \)) & Learned embedding for the item (\( p \)-th component-level embedding for \( x \)). \\ \hline
		\( \langle f(q), g(x) \rangle \) & Dot product similarity function: \( \langle f(q), g(x) \rangle = g(x)^T f(q) \). \\ \hline
	\end{tabular}
	\caption{Notation Table }
	\label{tab:Notations_table}
\end{table}

\subsection{Mixture of Logits}

The Mixture of Logits (MoL)\cite{zhai2023revisiting,ding2024efficient} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items. Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively. These embeddings are parameterized by neural networks that extract features from the query and item. The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting,ding2024efficient} are adaptive gating parameters constrained to sum to 1:

\begin{equation}
\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\end{equation}

This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.

To efficiently scale MoL for large datasets and hardware-optimized implementations, the formulation is extended by decomposing the dot products into batched outer products of query-side and document-side embeddings. This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm:\cite{zhai2023revisiting}

\begin{equation}
\phi(q,x) = \sum_{pq=1}^{P_q} \sum_{px=1}^{P_x} \pi_{pq,px}(q,x) \frac{\langle f_{pq}(q), g_{px}(x) \rangle}{|| f_{pq}(q) ||_2 || g_{px}(x) ||_2}
\end{equation}

Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications. Furthermore, it is possible to decompose any high-rank matrix into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.

@@

\noindent
1. \textbf{Component-Level Embeddings}:  
Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
\begin{equation}
X_p \leftarrow \{g_p(x) \mid x \in X\}.
\end{equation}

\noindent
2.\textbf{ Initial Candidate Retrieval:} This step involves computing similarity scores between a query representation and item representations for each feature component $p \in P$.

\begin{equation}
S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
\label{eq:initial_retrieval}
\end{equation}
Here, $f_p(q)$ and $g_p(x)$ represent the feature embeddings of the query $q$ and item $x$ for the $p$-th component, respectively. The dot product $\langle f_p(q), g_p(x) \rangle$ measures the relevance of each item $x \in X_p$ with respect to $q$, producing a set of scores $S_p$ .

@@

	\begin{algorithmic}[1]

		\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
		\For{each component $p \in P$}
		\State $X_p \gets \{g_p(x) \mid x \in X\}$ \Comment{Precompute embeddings}
		\EndFor

@@

		\textbf{4. Select Exact Top-k Items:}
		\For{each component $p \in P$}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
		\EndFor
		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\newpage
\subsection{Evaluation metrics in recommendation system}

The retrieval algorithm is integrated into a recommendation system, and its effectiveness is assessed using standard ranking metrics. These metrics are widely used in sequential recommendation tasks to evaluate the model's ability to rank relevant items higher in a user's preference list:

>>>>

% sage_latex_guidelines.tex V1.20, 14 January 2017

\documentclass[Afour,sageh,times]{sagej}

\usepackage{moreverb,url}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2016}

\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{subcaption}
\usepackage{booktabs} % For professional-quality rules
\usepackage{makecell} % For multiline headers
\begin{document}

\runninghead{Smith and Wittkopf}

\title{Boosting RAG Efficiency with dyRAG: Dynamic Candidate Selection for Optimal Retrieval}

\author{Zoubida Asmaa Boudjenane \affilnum{1} and Mohammed SALEM \affilnum{2}}

\affiliation{\affilnum{1}Computer Science Department, University of Mascara, Algeria\\
\affilnum{2}LISYS Laboratory, University of Mascara, Algeria}

\corrauth{Mohammed SALEM, LISYS Lab, University of Mascara, BP 305 Route Mamounia, Mascara, 29000, Algeria.}

\email{salem@univ-mascara.dz}

\begin{abstract}
Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge. However, its effectiveness heavily depends on the quality of the retrieved documents. Using a fixed number of retrieved documents K may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones. To address this issue, we propose dyRAG, a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency. Our method improves retrieval performance by maximizing relevant information and minimizing noise. Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems. Compared to traditional methods, it dynamically adjusts based on the query and dataset characteristics, offering greater adaptability across various contexts.
\end{abstract}

\keywords{Retrieval-Augmented Generation (RAG), Dynamic Retrieval (dyRAG), large Llanguage model, 	Query Complexity, Relevant Information}

\maketitle
\section{Introduction}
Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text \cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation \cite{Gu2018SearchEngine}, text generation \cite{liang2024controllabletextgenerationlarge} and question answering \cite{Zhang2023}. Notable examples of LLMs include OpenAI GPT series \cite{brown2020language}, Google BERT \cite{devlin2018bert} and T5 \cite{raffel2020exploring}. Despite their impressive capabilities, LLMs suffer from hallucination \cite{Radford2019}, often generating confident but incorrect information. This is a critical challenge for knowledge-intensive tasks requiring high accuracy \cite{Radford2019}. Additionally, their static nature being trained on fixed datasets \cite{Radford2019} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information \cite{Mousavi2025}.

Researchers have explored various approaches to address this limitation, including transfer learning \cite{zhuang2020comprehensive}, fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks. Despite their strengths, they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{Chuyuan2025}, which integrates real-time retrieval from external knowledge sources with generative models, enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations. For instance, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally, a critical challenge is determining the optimal number of retrieved passages K to feed into the generator \cite{lewis2020retrieval}. Often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes K a key factor in balancing performance and efficiency.

@@

In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL) \cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs. The use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches. Unlike traditional approaches, dyRAG adapts to the query and dataset characteristics, making it more flexible for diverse contexts. Furthermore, its design for handling large datasets ensures scalable and precise retrieval, which is crucial for RAG systems working with extensive corpora.

@@

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval \cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method, the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models \cite{Lafferty2001}, TF-IDF \cite{Robertson1997} and BM25 \cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents K is fixed, it fails to adapt to the varying complexity of queries. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.

Dynamic top-K Selection in Retrieval :addresses the limitations of static K by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC) \cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal K for retrieval, a reinforcement learning-based approach \cite{Zhou2021} where the agent learns to adjust K by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of K Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations. They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability. These systems might use a static K as a baseline but dynamically adjust it such as, the "Blended RAG" \cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, Furthermore STAYKATE (Static-Dynamic Hybrid Selection) \cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation) \cite{hei2024drragapplyingdynamicdocument} First, it retrieves an initial set of $k_1$ documents. Then, a classifier dynamically evaluates their relevance and retrieves additional relevant $K_2$ documents to enhance recall. However hybrid methods may face challenges in determining when to apply static or dynamic K-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt K to the unique characteristics of each query and dataset.

@@

\section{Overview of Retrieval-Augmented Generation }\label{sec3}
Retrieval-Augmented Generation (RAG) is a method designed to enhance the performance of large language models (LLMs) by integrating external, reliable knowledge sources into their response generation process \cite{awsRAG}. RAG combines search capabilities with LLM prompting. The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt \cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{rag.JPG}
	\caption{RAG system component \cite{gupta20241}. }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components: the retriever and the generator. The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows:
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}which relies on exact term matching techniques like BM25 and TF-IDF, dense retrieval\cite{karpukhin2020dpr}which Uses deep learning models to encode queries and documents into dense vector embeddings, retrieving relevant documents based on semantic similarity rather than exact word matching, and others. For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:
\begin{itemize}
\item Encoding each object into a specific representation (e.g., term-based for sparse retrieval, vector embeddings for dense retrieval).
\item Building an index to organize and efficiently retrieve relevant data during search .
\end{itemize}
This structured approach ensures that the most relevant information is retrieved quickly and accurately \cite{zhao2024retrievalaugmentedgenerationaigeneratedcontent}.

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator  \cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response.

@@

In various generative tasks, different models are selected based on the specific requirements: Models like BART \cite{lewis2019bart}, GPT \cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation, Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images, for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

@@

\begin{table}[h]
	\centering
	\large  % Increase overall font size
	\begin{tabular}{|c|p{5cm}|}  % Use p{width} for text wrapping
		\hline
		\textbf{Notation} & \textbf{Description} \\ \hline
		\( q \) (\( Q \), \( |Q| \)) & A single query (set of all queries, total number of queries). \\ \hline
		\( x \) (\( X \), \( |X| \)) & A single item (set of all items, total number of items). \\ \hline
		\( \phi(q, x) \) & The learned similarity function, Mixture-of-Logits (MoL), which computes the similarity between \( q \) and \( x \). \\ \hline
		\( P \) (\( P_q \), \( P_x \)) & Total number of low-rank embeddings (\( P = P_q \times P_x \)), where \( P_q \) and \( P_x \) are the number of embeddings for \( q \) and \( x \), respectively. \\ \hline
		\( \pi_p(q, x) \)  & Weight for the \( p \)-th (or \( p_q \)-th and \( p_x \)-th) embedding set for \( (q, x) \). \\ \hline
		\( f(q) \) (\( f_p(q) \)) & Learned embedding for the query (\( p \)-th component-level embedding for \( q \)). \\ \hline
		\( g(x) \) (\( g_p(x) \)) & Learned embedding for the item (\( p \)-th component-level embedding for \( x \)). \\ \hline
		\( \langle f(q), g(x) \rangle \) & Dot product similarity function: \( \langle f(q), g(x) \rangle = g(x)^T f(q) \). \\ \hline
	\end{tabular}
	\caption{Notation Table }
	\label{tab:Notations_table}
\end{table}
\subsection{Mixture of Logits} 
The Mixture of Logits (MoL) \cite{zhai2023revisiting,ding2024efficient} method provides a flexible and adaptive mechanism for computing similarity scores between queries and items. Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively. These embeddings are parameterized by neural networks that extract features from the query and item. The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q,x)\) \cite{zhai2023revisiting,ding2024efficient} are adaptive gating parameters constrained to sum to 1:
\begin{equation}
\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\end{equation}
This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.

To efficiently scale MoL for large datasets and hardware-optimized implementations, the formulation is extended by decomposing the dot products into batched outer products of query-side and document-side embeddings. This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm: \cite{zhai2023revisiting}

\begin{equation}
\phi(q,x) = \sum_{pq=1}^{P_q} \sum_{px=1}^{P_x} \pi_{pq,px}(q,x) \frac{\langle f_{pq}(q), g_{px}(x) \rangle}{|| f_{pq}(q) ||_2 || g_{px}(x) ||_2}
\end{equation}
Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications. Furthermore, it is possible to decompose any high-rank matrix into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.

@@

\noindent
1. \textbf{Component-Level Embeddings}:  
Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
\begin{equation}
X_p \leftarrow \{g_p(x) \mid x \in X\}.
\end{equation}
\noindent
2.\textbf{ Initial Candidate Retrieval:} This step involves computing similarity scores between a query representation and item representations for each feature component $p \in P$.  
\begin{equation}
S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
\label{eq:initial_retrieval}
\end{equation}
Here, $f_p(q)$ and $g_p(x)$ represent the feature embeddings of the query $q$ and item $x$ for the $p$-th component, respectively. The dot product $\langle f_p(q), g_p(x) \rangle$ measures the relevance of each item $x \in X_p$ with respect to $q$, producing a set of scores $S_p$ .

@@

	\begin{algorithmic}[1]
		\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
		\For{each component $p \in P$}
		\State $X_p \gets \{g_p(x) \mid x \in X\}$ \Comment{Precompute embeddings}
		\EndFor

@@

		\textbf{4. Select Exact Top-k Items:}
		\For{each component $p \in P$}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
		\EndFor
		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\subsection{Evaluation metrics in recommendation system}

The retrieval algorithm is integrated into a recommendation system, and its effectiveness is assessed using standard ranking metrics. These metrics are widely used in sequential recommendation tasks to evaluate the model's ability to rank relevant items higher in a user's preference list:
\begin{itemize}
	\item \textbf {Hit Rate at \( K \)} (HR@K) evaluates the proportion of users who have at least one relevant item in their top-\( K \) recommendations. It is computed as \cite{eviden2025mrr}:
